---
title: "p-likelihood: p-curve from a model comparison perspective"
author: "Felix Schönbrodt"
date: "Feb 26 2015"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
# library(rmarkdown); render("p-likelihood-report.Rmd") #
# library(rmarkdown); run("p-likelihood-report.Rmd")

library(shiny)
library(dplyr)
library(reshape2)
# load(file="res_10000.RData") #

load(file="res_5000_4.RData")

LR.boundary <- 5
res$LR_meta.decision <- "indecisive"
res$LR_meta.decision[res$LR_meta < 1/LR.boundary] <- "H0"
res$LR_meta.decision[res$LR_meta > LR.boundary] <- "H1"

res$LR_single.decision <- "indecisive"
res$LR_single.decision[res$LR_single < 1/LR.boundary] <- "H0"
res$LR_single.decision[res$LR_single > LR.boundary] <- "H1"

pcurve.boundary <- .05
res$pcurve.decision <- "indecisive"
res$pcurve.decision[res$p_lack < pcurve.boundary] <- "H0"
res$pcurve.decision[res$p_hack < pcurve.boundary] <- "H0"
res$pcurve.decision[res$p_evidence < pcurve.boundary] <- "H1"

# construct a long data frame of the decisions for tabulation
decisions.wide <- res %>% select(k, n, reality, LR_meta.decision, LR_single.decision, pcurve.decision)
decisions.long <- melt(decisions.wide, id.vars=c("k", "n", "reality"))

colnames(decisions.long) <- c("k", "n", "reality", "procedure", "decision")
levels(decisions.long$reality) <- c(
	"Reality: d=0 + intensive hacking", 
	"Reality: d=0.2 + intensive hacking", 
	"Reality: d = 0", 
	"Reality: d = 0.5"
)
levels(decisions.long$procedure) <- c("LR meta", "LR single", "p-curve")
```


# Method
## Simulation of three types of studies

- "H0": `r nrow(res)/3` simulated sets of k=5 studies with n=50 each (2 groups with n=25 each). Only keep sets where all 5 studies are significant. (So in all scenarios, the studies are more or less "p-hacked", as the non-significant results go to the file drawer).
- "H1": As H0, but with a true effect of d = 0.5
- "hack": As H1, but with a small true effect of d = 0.2. Sample size is optionally increased by 5 participants in each group, until either the maximum n of 100 is reached, or the study is significant. Again, only significant studies are reported.

## Computation of indices ##

- Classical p-curve (a la Simonsohn, Nelson, & Simmons, 2013; see (http://www.p-curve.com))
- Likelihood approach with a common meta-analyzed effect size across all p values ("LR_meta")
- Likelihood approach with a separat effect size for all p values ("LR_single"). That means, we assume that each p value gets it's own optimal p-curve, and that the empirical ES estimate is exactly the true ES.

# Results



#### Classification performance

If we take a LR bound of 5 (resp. 1/5), and a p-curve threshold of p < .05, we get:



```{r echo=FALSE, eval=FALSE}
with(decisions.long, prop.table(table(procedure, decision, reality, k, n), margin=c(1, 3:5))*100)

library(ggplot2)
library(dplyr)

ggplot(decisions.long, aes(x=procedure, fill=decision)) + geom_bar(stat="bin", position=position_stack()) + xlab("Procedure") + facet_wrap(reality~n) + coord_flip() + theme_bw() + scale_fill_manual(values=c("red3", "Chartreuse3", "grey60"))
```



Hence, classification performance is roughly in the same magnitude.

Some differences are there, though:

- LR_meta and curve are quite balanced concerning wrong decisions 
(3-5%), but LR_meta makes more correct decisions and less indecisive 
outcomes.
- LR_single is biased towards H1: It detects H1 quite good when it's 
there, but also assumes an effect if H0 is true with around 2x the error 
rate of the others.
- If this method should be used as a p-hacking-detector, researchers 
probably are most concerned about a wrong classification with a false 
alarm (i.e., reality = H1, decision = "no evidence"). For that purpose, 
LR_single seems to be a good candidate, as it has the lowest error rate 
of that kind, and the highest probability of detecting an effect if it 
really is there. (At the cost, as always, to misclassify some null 
effects as "evidence").

By tuning the LR bound one could roughly mimic the p-curve behavior (If 
this is of any interest ...).




# Discussion #


##

Although the LR approach is not dramatically better, I do see several advantages:

- p-likelihood has a natural, interpretable scale ("These results are four times more likely to come from a null effect, or even p-hacked set of studies, than from a real effect.")
- p-likelihood is a continuous measure of evidence. Even if a result does not cross the arbitrary bound of 5, the LR provides a continuous measure of evidence - in contrast to the binary decision provided by p-curve's chi2-tests.
- Boundaries can be fine tuned (would be somewhat)
- If a classification is requested, one has less non-decisions than with p-curve, and a comparable or better classification quality.

A crucial assumption (and feature!) of classical p-curve is that it assumes that the selection pressure for p-values <= .05 and >.05 is different. Therefore it also cannot easily be extended to a range of p values > 5%, as probably most of these p values drop out of the literature.

## Differences of approaches

- p-curve only takes *p* values into account. p-likelihood additionally takes effect size into account.
- p-curve has a three-way classification ("evidence", "lack of evidence", "hacking"), p-likelihood only a two-way ("more consistent with H0", "more consistent with H1"). However, separating null effects from p-hacked effects would require to define a specific model for the distribution of hacked *p* values. Given the many creative ways of p-hacking, it seems quite difficult to obtain a defensible model.
- In the presence of a real effect, both methods could come to different conclusions, as they test different models. Consider the following situation: A set of studies claims an enormous effect of d=1.5, but the p-values indicate a much smaller, but existing effect. p-curve only tests for the skew of the p-curve, and if the small effect is in fact present, p-curve will indicate "evidential value". Hence, it tests for the presence of *any* effect. p-likelihood, in contrast, looks whether the p values are consistent with the effect size *that is claimed in the paper*. If the p-values are consistent with a small effect, but inconsistent with the claimed huge effect, p-likelihood could indicate that the H0 is more consistent than H1.
- 	TODO: A plot of the situation
- Both approaches can make sense. If one asks, whether there is any evidential value (regardless of the magnitude of the effect), p-urve gives the better answer. If onw aks whether the p-values support the effect that the authors claim, p-likelihood gives the better answer.
- Problem p-curve: The typical problem "with enough n, everything gets significant": If enough studies are included, any trivial skewness will indicate "evidential value". This can lead to the situation where p-curve indicates both "lack of evidence" (i.e., empirical p-curve is flatter than 33% power curve) *and* evidential value (because there is a significant skew).

## A potential Bayesian extension

A debatable assumption of p-likelihood is that the true effect size for which the p-curve is constructed is fixed and known. If a prior is put on the H1 model the evidence ratio would be a Bayes factor, which takes uncertainty about the true effect size into account.


```{r, echo=FALSE}
inputPanel(
	sliderInput("LR_bound", label = "LR boundary:", min = 1, max = 10, value = 5, step = 0.1),
	sliderInput("p_bound", label = "p-curve - critical p value:", min = 0, max = 0.2, value = .05, step = 0.005),
	selectInput("reality", label = "Generating model:", c("H0"="H0", "H1: d = 0.5"="H1", "Hack: d=0, p-hacked"="hack0", "Hack: d=0.2, p-hacked"="hack0.2"))
)

renderPlot({
	# input <- list(reality="hack0", LR_bound=5, p_bound=.05)
	res0 <- filter(res, reality==input$reality)
	
	LR.boundary <- input$LR_bound
	res0$LR_meta.decision <- "indecisive"
	res0$LR_meta.decision[res0$LR_meta < 1/LR.boundary] <- "H0"
	res0$LR_meta.decision[res0$LR_meta > LR.boundary] <- "H1"
	
	res0$LR_single.decision <- "indecisive"
	res0$LR_single.decision[res0$LR_single < 1/LR.boundary] <- "H0"
	res0$LR_single.decision[res0$LR_single > LR.boundary] <- "H1"
	
	pcurve.boundary <- input$p_bound
	res0$pcurve.decision <- "indecisive"
	res0$pcurve.decision[res0$p_lack < pcurve.boundary] <- "H0"
	res0$pcurve.decision[res0$p_hack < pcurve.boundary] <- "H0"
	res0$pcurve.decision[res0$p_evidence < pcurve.boundary] <- "H1"
	
	# construct a long data frame of the decisions for tabulation
	decisions.wide <- res0 %>% select(k, n, reality, LR_meta.decision, LR_single.decision, pcurve.decision)
	decisions.long <- melt(decisions.wide, id.vars=c("k", "n", "reality"))
	
	colnames(decisions.long) <- c("k", "n", "reality", "procedure", "decision")
	levels(decisions.long$reality) <- c("Reality: d=0 + intensive hacking", "Reality: d=0.2 + intensive hacking", "Reality: d = 0", "Reality: d = 0.5")
	levels(decisions.long$procedure) <- c("LR meta", "LR single", "p-curve")
	
	# summarise to get percentages
	d3 <- decisions.long %>% group_by(procedure, n, k, decision) %>% summarise(N=n()) %>% mutate(perc=N/sum(N))
	
	ggplot(d3, aes(x=procedure, y=perc, fill=decision)) + geom_bar(stat="identity", position=position_stack()) + xlab("Procedure") + facet_wrap(n~k) + coord_flip() + theme_bw() + scale_fill_manual(values=c("red3", "Chartreuse3", "grey60"))
})
```


```{r echo=FALSE, eval=FALSE}
# ROCR #

library(ROCR)

d <- decisions.wide
d$reality <- as.character(d$reality)
d$reality[d$reality=="hack"] <- "H0"

d2 <- filter(d, LR_meta.decision != "indecisive", n==40, k==5)
pr <- prediction(d2$LR_meta.decision, d2$reality)

pr <- prediction(factor(c("H0", "H0", "H0", "H1")), factor(c("H0", "H0", "H1", "H1")))


d <- res
d$reality <- as.character(d$reality)
d$reality[d$reality=="hack"] <- "H0"

d$pred <- log(d$LR_meta)

d2 <- filter(d, n==40, k==5)

# classical ROC
pr <- prediction(d2$pred, d2$reality)
perf <- performance(pr, measure="tpr", x.measure="fpr")
plot(perf)

# Accuracy is optimal with cutoff at LR=1 (well, that's obvious ...)
perf2 <- performance(pr, measure="acc")
plot(perf2)

perf3 <- performance(pr, measure="sens", x.measure="spec")
plot(perf3)


```

